# Community composition

```{r load_data_community}
load("resources/data.Rdata")
```

## Taxonomy overview 

### Stacked barplot

```{r taxonomy_barplot, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  mutate(individual=factor(individual,levels=c("Sg1","Sg2","Sg3","Sg4","Sg5","Sg6","Sg7","Sg8","Sg9","Sg10"))) %>%
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    facet_nested(. ~ individual,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

### Phylum relative abundances
```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum) %>%
  summarise(relabun=sum(count))

phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    arrange(-mean) %>%
    tt()
```

```{r taxonomy_boxplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=mean(relabun)) %>%
    arrange(-mean) %>%
    select(phylum) %>%
    pull()

phylum_summary %>%
    filter(phylum %in% phylum_arrange) %>%
    mutate(phylum=factor(phylum,levels=rev(phylum_arrange))) %>%
    ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum)) +
        scale_color_manual(values=phylum_colors[rev(phylum_arrange)]) +
        geom_jitter(alpha=0.5) + 
        theme_minimal() + 
        theme(legend.position="none") +
        labs(y="Phylum",x="Relative abundance")
```

## Taxonomy boxplot

### Family
```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family) %>%
  summarise(relabun=sum(count))

family_summary %>%
    group_by(family) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    arrange(-mean) %>%
    tt()
```


```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    filter(family != "f__")%>%
    arrange(-mean) %>%
    select(family) %>%
    mutate(family=sub("^f__", "", family)) %>%
    pull()

family_summary %>%
    left_join(genome_metadata %>% 
                select(family,phylum) %>% 
                unique(),
              by=join_by(family==family)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
      filter(family != "f__")%>%
    mutate(family=sub("^f__", "", family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~type)+
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")

```

### Genus
```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,genus) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__")

genus_summary %>%
    group_by(genus) %>%
    summarise(mean=mean(relabun),sd=sd(relabun)) %>%
    arrange(-mean) %>%
    tt()
```

```{r taxonomy_jitterplot_genus, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary %>%
    left_join(genome_metadata %>% select(genus,phylum) %>% unique(),by=join_by(genus==genus)) %>%
    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    filter(genus %in% genus_arrange[1:20]) %>%
    mutate(genus=factor(genus,levels=rev(genus_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=genus, group=genus, color=phylum)) +
        scale_color_manual(values=phylum_colors[-c(3,4,6,8)]) +
        geom_jitter(alpha=0.5) + 
        facet_grid(.~type)+
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")

```

## Alpha diversity

```{r alpha_div, comment="", message=FALSE, warning=FALSE}
# Calculate Hill numbers
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))
```

```{r alpha_div_plot, comment="", message=FALSE, warning=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>%
      ggplot(aes(y = value, x = type, group=type, color=type, fill=type)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(name="Sample type",
          breaks=c("cloaca","feces"),
          labels=c("Cloaca","Faeces"),
          values=c("#e5bd5b", "#6b7398")) +
      scale_fill_manual(name="Sample type",
          breaks=c("cloaca","feces"),
          labels=c("Cloaca","Faeces"),
          values=c("#e5bd5b50", "#6b739850")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)
      )
```

```{r alpha_div_test, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  lmerTest::lmer(richness ~ type + (1 | individual), data = ., REML = FALSE) %>%
  broom.mixed::tidy() %>%
  tt()

alpha_div %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  lmerTest::lmer(neutral ~ type + (1 | individual), data = ., REML = FALSE) %>%
  broom.mixed::tidy() %>%
  tt()

alpha_div %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  lmerTest::lmer(phylogenetic ~ type + (1 | individual), data = ., REML = FALSE) %>%
  broom.mixed::tidy() %>%
  tt()

alpha_div %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  lmerTest::lmer(functional ~ type + (1 | individual), data = ., REML = FALSE) %>%
  broom.mixed::tidy() %>%
  tt()
```

## Beta diversity

```{r beta_div, comment="", message=FALSE, warning=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  hillpair(., q = 1, dist = dist)
```

```{r permanova, comment="", message=FALSE, warning=FALSE}
#Richness
betadisper(beta_q0n$C, sample_metadata$type) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q0n$C ~ type, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()

#Neutral diversity
betadisper(beta_q1n$C, sample_metadata$type) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q1n$C ~ type, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()

#Phylogenetic diversity
betadisper(beta_q1p$C, sample_metadata$type) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q1p$C ~ type, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()

#Functional diversity
betadisper(beta_q1f$C, sample_metadata$type) %>% permutest(., pairwise = TRUE) 
adonis2(beta_q1f$C ~ type, 
        data = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))), 
        permutations = 999, 
        strata = sample_metadata %>% arrange(match(sample,labels(beta_q1n$C))) %>% pull(individual)) %>%
        broom::tidy() %>%
        tt()
```


### Richness diversity plot

```{r beta_div_nmds_richness_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  mutate(individual=factor(individual, levels=c("Sg1","Sg2","Sg3","Sg4","Sg5","Sg6","Sg7","Sg8","Sg9","Sg10"))) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = type, shape = individual)) +
    scale_color_manual(name="Sample type",
          breaks=c("cloaca","feces"),
          labels=c("Cloaca","Faeces"),
          values=c("#e5bd5b", "#6b7398")) +
    scale_shape_manual(values = 1:10) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="Individual")
```

### Neutral diversity plot

```{r beta_div_nmds_neutral_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  mutate(individual=factor(individual, levels=c("Sg1","Sg2","Sg3","Sg4","Sg5","Sg6","Sg7","Sg8","Sg9","Sg10"))) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = type, shape = individual)) +
    scale_color_manual(name="Sample type",
          breaks=c("cloaca","feces"),
          labels=c("Cloaca","Faeces"),
          values=c("#e5bd5b", "#6b7398")) +
    scale_shape_manual(values = 1:10) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="Individual")
```

### Phylogenetic diversity plot

```{r beta_div_nmds_phylo_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1p$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  mutate(individual=factor(individual, levels=c("Sg1","Sg2","Sg3","Sg4","Sg5","Sg6","Sg7","Sg8","Sg9","Sg10"))) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = type, shape = individual)) +
    scale_color_manual(name="Sample type",
          breaks=c("cloaca","feces"),
          labels=c("Cloaca","Faeces"),
          values=c("#e5bd5b", "#6b7398")) +
    scale_shape_manual(values = 1:10) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(shape="Individual")
```

The figure is distorted due to the total phylogenetic turnover (100% a single Pseudomonadota MAG vs. 100% a single Campylobacterota MAG) between some of the cloacal communities.

### Functional diversity plot

```{r beta_div_nmds_functional_plot, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
beta_q1f$C %>%
  vegan::metaMDS(., trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(individual=factor(individual, levels=c("Sg1","Sg2","Sg3","Sg4","Sg5","Sg6","Sg7","Sg8","Sg9","Sg10"))) %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = type)) +
    scale_color_manual(name="Sample type",
          breaks=c("cloaca","feces"),
          labels=c("Cloaca","Faeces"),
          values=c("#e5bd5b", "#6b7398")) +
    scale_shape_manual(values = 1:10) +
    geom_point(size = 4) +
    #   stat_ellipse(aes(color = beta_q1n_nmds$Groups))+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    )
```